Linear Variation and an Optimization of Algorithms for Connected Components Labeling in a Binary Image

### 2016-04-06 paper/paper.tex
Add russian translation of key sections

@@

@@

We will use the fact that for the case of 8-connectivity all $4$ pixels of any
$2 \times 2$ square are adjacent to each other, and so are contained in the same CC
and should have same labels upon completion of the labeling algorithm.
So instead of labeling single pixels we can label $2 \times 2$ groups of pixels,
or ``cells''.

Consider a $2N \times 2M$ binary image\footnote{
For convenience in this paper we consider only input images with even size
in both dimensions. If this is not the case, one can easily append a row or
a column of background pixels to the image.}
specified by a binary predicate $b(x, y)$,
returning $1$ in case the pixel at the given coordinates is black (object pixel),
and $0$ otherwise (background pixel).
For convenience we use 0-indexation throughout the paper, so the topmost row has
row index $y=0$, and the leftmost column has column index $x=0$.

The condensed image will be of size $N \times M$.
Each pixel of the condensed image will unambiguously represent the configuration
of the corresponding $2\times 2$ cell of pixels of the initial binary image.
So the condensed image will contain pixels of $2^4 = 16$ colors.
We suggest the function $c(x, y)$ denoting the color of a pixel of condensed image
to be the following:
\begin{align*}
  c(x, y) &= 2^0 b(2x, 2y)   + 2^1 b(2x+1, 2y) \\
          &+ {} 2^2 b(2x, 2y+1) + 2^3 b(2x+1,2y+1)
\end{align*}

So the value of each pixel of the initial image is stored in the corresponding
bit in color of the corresponding pixel of the condensed image.
Note that if $c(x, y) = 0$ for some $(x, y)$, then there are no object pixels in
this cell, and no label should be assigned to this whole cell. The enumeration of
pixels inside one cell is shown on \autoref{fig:pixelsEnumeration}.

@@

We need also to define connectivity function for the condensed image.
We cannot actually just use 8-connectivity, as whether two cells are adjacent
cells of one CC depends not only on their coordinates, but also on their
configuration. See \autoref{fig:connectivity:examples} for examples.

@@

However, for every variant of relative placement of two nearby cells there is
a mask of pixels of each cells that are important for direct connectivity of
this pair of cells.
So two cells are considered directly connected, if they are nearby
in terms of 8-connectivity and each of them contains at least one object pixel
in the mask of important pixels corresponding to relative placement of the cells.
The masks are shown in \autoref{fig:connectivity:masks}.

@@

The idea is that if we naturally encode mask pixels as 4-bit numbers in the
way similar to how we defined $c$, we can express the predicate denoting whether
two nearby cells
are directly connected efficiently with just two bitwise and one logical AND
operations.
As an example, \autoref{fig:connectivity:code} compares possible codes in \C checking
if the labels of the current pixel or cell and the one right on top of it
should be same for classic 8-connectivity and for the new connectivity.
Note that as bitwise operations are usually relatively cheap, we added only a
little complication compared to decreasing the number of pixels to process by the 
factor of 4.

@@

As each of these methods, omitting DTSUF could be used with both condensed and
non-condensed images, for convenience we denote height and width of the image
on input of each algorithm as $H$ and $W$.

Also to avoid repeating ``pixel or cell'' we will just use the word ``pixel''
meaning the appropriate raster for non-condensed and condensed images.
When we mention neighbors of some pixel, we mean pixels that are directly connected
to it in terms of appropriate connectivity.

@@

Another general approach for graph CC labeling which is more popular in image
labeling is using some kind of a data structure to maintain disjoint sets of pixels
that are considered to be in the same CC.

The structure is commonly called Union-Find, as the two essential operations
it is supposed to support are Union, which is used to merge some two of the
disjoint sets in one, and Find, which is used to find the representative element
of the set some element belongs to, so that we can always tell if two elements are
in the same set.
The structure can be implemented so that the amortized time complexity for each
operation is $O(\alpha(HW))$, where $\alpha$ is the inverse Ackermann function,
and $\alpha(HW) < 5$ for all remotely practical image sizes\cite{CLRS}. % TODO links

The approach that is named SUF (Scan plus Union-Find) in this paper follows.

Pixels are processed in the natural order: from top to bottom and from
left to right, so each row is fully processed before all rows after it.
So for each object pixel we encounter there are at most 4 nearby pixels
that were already processed, see \autoref{fig:suf:neighbors}.

At first each object pixel is considered to belong to its own CC, so right before
considering the neighbors we increment the CC counter.
Moreover, it is associated with its own disjoint set in the Union-Find structure.
More specifically, object pixel at position $x, y$ is associated with
the set number $xW + y$ to guarantee absence of collisions.

The possible neighbors are processed in the order shown on \autoref{fig:suf:neighbors},
and for each neighbor that is object pixel, the Union operation is performed
on the two sets containing the current and the neighbor pixel.
In case the Union operation was successful, which means that the pixels
had belonged previously to different sets and we have just found a way to merge two
CC, the CC counter is decremented.

@@

Note that this algorithm outputs not the matrix of final labels, but only the number
of CC.
This is the place in which it differs from some common two-pass approaches for CC % TODO references
labeling, as it is an algorithm not for labeling, but just for counting CC.
The latter is exactly what is needed for the aforementioned linear variation problem.

Also we have to mention that for performance reasons the Union-Find internal structures
are allocated ahead to have the size of at least $HW$ to guarantee that
each object can be easily associated with unique starting set given only its
coordinates. So memory consumption of SUF is linear in the number of pixels.

@@

This approach is a modification of SUF designed to reduce memory consumption
and potential number of cache-misses generated by Union-Find operations.

The idea is to run through each row twice: first time to establish label equivalence
classes within row, and the second time to set unique equal label for all pixels of
same CC.
This gives us the opportunity to maintain just $O(W)$
different labels at each moment, and so the Union-Find structure can be allocated
for only $O(W)$ elements, as opposed to $WH$ elements needed for SUF.

The exact maximum number of label equivalence classes between two adjacent rows
depends on whether the given image is condensed.
For non-condensed images $W=2M$, and maximum number of label equivalence classes
in two adjacent rows is $M=\frac{W}2$, as if two object pixels are both in adjacent
rows and in adjacent columns, then they are neighbors in terms of 8-connectivity.

On the other hand, for condensed images $W=M$, and maximum number of label equivalence
class in two adjacent rows of cells is $2W=2M$.
See \autoref{fig:suf2:extremal} for illustration on the extremal cases.

@@

Our benchmarks show that, although SUF2 without condensation is generally more
efficient on noise images than SUF without condensation, the mentioned
consideration make the $2 \times 2$ optimization
of SUF2 less dramatic in means of performance increase factor.
Optimized SUF2 significantly outperforms optimized SUF only on
huge noise images with less than 60\% density, while falling behind on
small and medium-sized noise images.

Note that, while SUF2 is intended for CC counting, its memory
consumption is only $O(W)$, which is significant for embedded systems and
other resource-tight applications.

@@

This modification of SUF was implemented to compare the optimization we present with
another optimization suggested by Wu, Otoo and Suzuki\cite{wuotoo},
namely decision tree.
Unfortunately DTSUF (Decision Tree Scan plus Union-Find) is not
applicable for condensed image, as it exploits similar observations on 8-connectivity.

As before, the goal is to reduce the number of pixel lookup and label copy operations.
This is achieved with careful consideration of only the necessary neighbors,
stopping once the set of one or two label operations that are needed is
determined. The decision tree for processing an object pixel is
shown on \autoref{fig:dtsuf}.

>>>>

\usepackage[russian,english]{babel}

@@

% \newcommand{\texten}[1]{}
% \newcommand{\textru}[1]{\foreignlanguage{russian}{\emph{#1}}{}}

\newcommand{\texten}[1]{#1}
\newcommand{\textru}[1]{\\
  \foreignlanguage{russian}{\emph{#1}}}

@@

\texten{
We will use the fact that for the case of 8-connectivity all $4$ pixels of any
$2 \times 2$ square are adjacent to each other, and so are contained in the same CC
and should have same labels upon completion of the labeling algorithm.
So instead of labeling single pixels we can label $2 \times 2$ groups of pixels,
or ``cells''.
}\textru{
Наблюдение: с 8-связностью в любом квадратике $2 \times 2$ любые два пикселя
находятся рядом. Значит если это пара объектных пикселей, то они находятся в одной
КС, и значит каждый такой квадратик пересекается не более чем с одной КС.
Поэтому вместо маркировки отдельных пикселей можно маркировать такие квадраты.
Назовём их клетками.
}

\texten{
Consider a $2N \times 2M$ binary image\footnote{
  For convenience in this paper we consider only input images with even size
  in both dimensions. If this is not the case, one can easily append a row or
  a column of background pixels to the image.
}
specified by a binary predicate $b(x, y)$,
returning $1$ in case the pixel at the given coordinates is black (object pixel),
and $0$ otherwise (background pixel).
For convenience we use 0-indexation throughout the paper, so the topmost row has
row index $y=0$, and the leftmost column has column index $x=0$.
}\textru{
Рассмотрим бинарное изображение размера $2N \times 2M$\footnote{
  Для удобства мы рассматриваем только изображения с чётной высотой и шириной.
  В общем случае при необходимости можно добавить строку или столбец с фоновыми
  пикселями.
},
заданное предикатом $b(x, y)$, который возвращает $1$, если переданы координаты
объектного пикселя, и $0$ иначе.
Мы везде будем использовать $0$-индексацию: верхняя строка и левый столбец имеют
номер $0$.
}

\texten{
The condensed image will be of size $N \times M$.
Each pixel of the condensed image will unambiguously represent the configuration
of the corresponding $2\times 2$ cell of pixels of the initial binary image.
So the condensed image will contain pixels of $2^4 = 16$ colors.
We suggest the function $c(x, y)$ denoting the color of a pixel of condensed image
to be the following:
}\textru{
Сжатое изображение будет иметь размер $N \times M$.
Каждый пиксель полученного изображения должен будет однозначно задавать конфигурацию
соответствующих четырёх пикселей исходного изображения, поэтому он может быть одного
из $2^4 = 16$ цветов. Предлагается такая функция $c(x,y)$, кодирующая четыре исходных
пикселя в цвет клетки:
}
\begin{align*}
  c(x, y) &= 2^0 b(2x, 2y)   + 2^1 b(2x+1, 2y) \\
          &+ {} 2^2 b(2x, 2y+1) + 2^3 b(2x+1,2y+1)
\end{align*}

\texten{
So the value of each pixel of the initial image is stored in the corresponding
bit in color of the corresponding pixel of the condensed image.
Note that if $c(x, y) = 0$ for some $(x, y)$, then there are no object pixels in
this cell, and no label should be assigned to this whole cell. The enumeration of
pixels inside one cell is shown on \autoref{fig:pixelsEnumeration}.
}\textru{
Таким образом значение каждого из исходных пикселей просто записано в соответствующий
бит в итоговом пикселе. Важно отметить, что если все четыре исходных пикселя были
фоновые, то и клетка будет иметь цвет 0. Нумерация пикселей внутри клетки показана на
\autoref{fig:pixelsEnumeration}.
}

@@

\texten{
We need also to define connectivity function for the condensed image.
We cannot actually just use 8-connectivity, as whether two cells are adjacent
cells of one CC depends not only on their coordinates, but also on their
configuration. See \autoref{fig:connectivity:examples} for examples.
}\textru{
Ещё нужно дать определение связности для сжатого изображения. Мы уже не можем
использовать 8-связность: теперь нельзя сказать, связаны ли две клетки, не
проанализировав их цвета. Смотри \autoref{fig:connectivity:examples}.
}

@@

\texten{
However, for every variant of relative placement of two nearby cells there is
a mask of pixels of each cells that are important for direct connectivity of
this pair of cells.
So two cells are considered directly connected, if they are nearby
in terms of 8-connectivity and each of them contains at least one object pixel
in the mask of important pixels corresponding to relative placement of the cells.
The masks are shown in \autoref{fig:connectivity:masks}.
}\textru{
Тем не менее, можно заметить, что анализировать можно не все биты цвета, а только
некоторую маску тех бит, которые важны для связности двух конкретных клеток.
Эти маски показаны на \autoref{fig:connectivity:masks}.
}

@@

\texten{
The idea is that if we naturally encode mask pixels as 4-bit numbers in the
way similar to how we defined $c$, we can express the predicate denoting whether
two nearby cells
are directly connected efficiently with just two bitwise and one logical AND
operations.
As an example, \autoref{fig:connectivity:code} compares possible codes in \C checking
if the labels of the current pixel or cell and the one right on top of it
should be same for classic 8-connectivity and for the new connectivity.
Note that as bitwise operations are usually relatively cheap, we added only a
little complication compared to decreasing the number of pixels to process by the 
factor of 4.
}\textru{
Идея в том, что с нашим естественным способом кодирования конфигурации клеток в
4-разрядных числах, предикат, определяющий связность двух соседных клеток,
может быть представлен всего двумя битовыми и одной логической операцией И.
В качестве примера, на \autoref{fig:connectivity:code} приведено сравнение
возможных фрагментов программы на языке \C\hspace{-0.8em}, проверяющих, связан
ли текущий пиксель (или клетка) с пикселем (или клеткой) над ним (ней).
Стоит отметить, что хотя конструкция
усложнилась, битовые операции как правило очень эффективны, и это усложнение, скорее
всего, незначительно по сравнению с уменьшением количества рассматриваемых позиций в 4
раза.
}

@@

\texten{
As each of these methods, omitting DTSUF could be used with both condensed and
non-condensed images, for convenience we denote height and width of the image
on input of each algorithm as $H$ and $W$.
}\textru{
Так как все рассматриваемые методы, кроме DTSUF, можно применять как для сжатого,
так и для несжатого изображения, мы обозначим размеры входного изображения
каждого изображения за $H$ и $W$.
}

\texten{
Also to avoid repeating ``pixel or cell'' we will just use the word ``pixel''
meaning the appropriate raster for non-condensed and condensed images.
When we mention neighbors of some pixel, we mean pixels that are directly connected
to it in terms of appropriate connectivity.
}\textru{
По аналогии, чтобы избежать повторения ``пиксель или клетка'' мы будет просто
использовать слово ``пиксель'', имея в виду растр входного изображения.
Под соседями пикселя будем понимать пиксели, прямо связанные с ним в терминах
соответствующей связности.
}

@@

\texten{
Another general approach for graph CC labeling which is more popular in image
labeling is using some kind of a data structure to maintain disjoint sets of pixels
that are considered to be in the same CC.
}\textru{
Другой, более популярный в этом контексте общий подход к подсчёту компонент
связности состоит в том, чтобы использовать структуру данных для поддержания
набора непересекающихся множеств пикселей, чем, по сути, КС и являются.
}

\texten{
The structure is commonly called Union-Find, as the two essential operations
it is supposed to support are Union, which is used to merge some two of the
disjoint sets in one, and Find, which is used to find the representative element
of the set some element belongs to, so that we can always tell if two elements are
in the same set.
The structure can be implemented so that the amortized time complexity for each
operation is $O(\alpha(HW))$, where $\alpha$ is the inverse Ackermann function,
and $\alpha(HW) < 5$ for all remotely practical image sizes\cite{CLRS}. % TODO links
}\textru{
Эту структуру принято называть Union-Find, так как две основные операции,
которые она поддерживает, это Union~--- объединение двух множеств в одно, и
Find~--- нахождение характерного представителя для множества, содержащего данный
элемент.
Этот представитель может быть использован для проверки принадлежности двух
элементов одному и тому же множеству.
Эта структура данных хорошо изучена и может быть реализована так, что
амортизированная сложность каждой операции~--- всего $O(\alpha(HW))$, где
$\alpha$~--- обратная функция Аккермана.
Для всех практически возможных размеров изображения эта функция не превосходит 5.
}

\texten{
The approach that is named SUF (Scan with Union-Find) in this paper follows.
}\textru{
Далее следует описание подхода, который мы называем SUF (Scan with Union-Find).
}

\texten{
Pixels are processed in the natural order: from top to bottom and from
left to right, so each row is fully processed before all rows after it.
So for each object pixel we encounter there are at most 4 nearby pixels
that were already processed, see \autoref{fig:suf:neighbors}.
}\textru{
Пиксели обрабатываются в естественном порядке сверху вниз слева направо, так что
каждая строка начинает обрабатываться, когда все предыдущие строки уже
обработаны. Таким образом, у очередного пикселя может быть до четырёх пикселей,
которые находятся рядом и уже были обработаны.
}

\texten{
At first each object pixel is considered to belong to its own CC, so right before
considering the neighbors we increment the CC counter.
Moreover, it is associated with its own disjoint set in the Union-Find structure.
More specifically, object pixel at position $x, y$ is associated with
the set number $xW + y$ to guarantee absence of collisions.
}\textru{
Каждый объектный пиксель сначала предполагается принадлежащим своей КС.
Поэтому прежде чем начинать рассматривать соседей, мы увеличиваем счётчик.
Более того того, каждый пиксель изначально ассоциирован со своим собственным
независимым множеством в структуре Union-Find, состоящем из одного элемента.
Конкретно пиксель в позиции $x, y$ заранее ассоциирован с элементом номер $xW + y$,
чтобы гарантировать отсутствие коллизий.
}

\texten{
The possible neighbors are processed in the order shown on \autoref{fig:suf:neighbors},
and for each neighbor that is object pixel, the Union operation is performed
on the two sets containing the current and the neighbor pixel.
In case the Union operation was successful, which means that the pixels
had belonged previously to different sets and we have just found a way to merge two
CC, the CC counter is decremented.
}\textru{
Возможные соседи перебираются в порядке, показанном на
\autoref{fig:suf:neighbors}.
Для каждого объектного пикселя, который напрямую соседствует с текущим, мы
применяем операцию Union для множества, содержащих текущий пиксель и
множества, содержащего этого соседа.
Если эти множества не совпадали, то мы нашли место, где КС соединяются.
В этом случае уменьшаем счётчик на 1.
}

@@

\texten{
Note that this algorithm outputs not the matrix of final labels, but only the number
of CC.
This is the place in which it differs from some common two-pass approaches for CC % TODO references
labeling, as it is an algorithm not for labeling, but just for counting CC.
The latter is exactly what is needed for the aforementioned linear variation problem.
}\textru{
Отметим, что этот алгоритм выдаёт не матрицу финальных меток, а только
количество компонент связности.
Именно в этом месте он отличается от некоторых известных подходов к CC labeling.
Для текущей задачи нам нужно именно считать компоненты связности.
}

\texten{
Also we have to mention that for performance reasons the Union-Find internal structures
are allocated ahead to have the size of at least $HW$ to guarantee that
each object can be easily associated with unique starting set given only its
coordinates. So memory consumption of SUF is linear in the number of pixels.
}\textru{
Также необходимо отметить, что для лучшей производительности служебные массивы
структуры Union-Find заранее выделяются размера как минимум $HW$, чтобы
обеспечить простоту соответствия между пикселями изображения и элементами в
системе множеств.
Таким образом, SUF использует линейное от числа пикселей количество памяти
$O(HW)$.
}

@@

\texten{
This approach is a modification of SUF designed to reduce memory consumption
and potential number of cache-misses generated by Union-Find operations.
}\textru{
Этот подход~--- модификация SUF, предложенная, чтобы сократить потребление
памяти и предположительно сильно уменьшить число кэш-промахов, происходящих
из-за произвольного доступа к большим массивам структуры Union-Find.
}

\texten{
The idea is to run through each row twice: first time to establish label equivalence
classes within row, and the second time to set unique equal label for all pixels of
same CC.
This gives us the opportunity to maintain just $O(W)$
different labels at each moment, and so the Union-Find structure can be allocated
for only $O(W)$ elements, as opposed to $WH$ elements needed for SUF.
}\textru{
Идея в том, чтобы бежать по каждой строке дважды: один раз чтобы установить
классы эквивалентности меток в текущей и предыдущей строках, и второй раз, чтобы
расставить уникальные метки в пиксели каждого класса.
Это даёт нам возможность иметь не более $O(W)$ различных меток в каждый момент
времени, и поэтому Union-Find можно выделять размера всего $O(W)$, при том что в
SUF выделялось ровно $WH$ элементов.
}

\texten{
The exact maximum number of label equivalence classes between two adjacent rows
depends on whether the given image is condensed.
For non-condensed images $W=2M$, and maximum number of label equivalence classes
in two adjacent rows is $M=\frac{W}2$, as if two object pixels are both in adjacent
rows and in adjacent columns, then they are neighbors in terms of 8-connectivity.
}\textru{
Точная оценка на максимальное число классов эквивалентности меток в двух
соседних строках зависит от того, сжатое изображение поступило на вход, или нет.
Для несжатых изображений $W=2M$, и наибольшее число классов эквивалентности в
соседних строках равно $M=\frac{W}2$, так как если два объектных пикселя
находятся одновременно в соседних строках и соседних столбцах, то они соседи в
смысле 8-связности.
}

\texten{
On the other hand, for condensed images $W=M$, and maximum number of label equivalence
class in two adjacent rows of cells is $2W=2M$.
See \autoref{fig:suf2:extremal} for illustration on the extremal cases.
}\textru{
С другой стороны, для сжатых изображений $W=M$, и максимальное число классов
эквивалентности равно $2W=2M$. Смотри примеры на \autoref{fig:suf2:extremal}.
}

@@

\texten{
Our benchmarks show that, although SUF2 without condensation is generally more
efficient on noise images than SUF without condensation, the mentioned
consideration make the $2 \times 2$ optimization
of SUF2 less dramatic in means of performance increase factor.
Optimized SUF2 significantly outperforms optimized SUF only on
huge noise images with less than 60\% density, while falling behind on
small and medium-sized noise images.
}\textru{
Проведённые эксперименты показывают, что хотя SUF2 без сжатия в целом более
эффективен, чем SUF без сжатия, приведённая оценка делает относительное
ускорение SUF2 со сжатием не таким значительным.
}

\texten{
Note that, while SUF2 is intended for CC counting, its memory
consumption is only $O(W)$, which is significant for embedded systems and
other resource-tight applications.
}\textru{
Отметим, что SUF2 решает задачу подсчёта СК, используя всего $O(W)$
дополнительной памяти.
}

@@

\texten{
This modification of SUF was implemented to compare the optimization we present with
another optimization suggested by Wu, Otoo and Suzuki\cite{wuotoo},
namely decision tree.
Unfortunately DTSUF (Decision Tree Scan plus Union-Find) is not
applicable for condensed image, as it exploits similar observations on 8-connectivity.
}\textru{
Эта модификация SUF была добавлена чтобы сравнить нашу оптимизацию с
одной из оптимизаций, предложенных Wu, Otoo и Suzuki\cite{wuotoo}, а именно с
деревом принятия решения.
К сожалению, DTSUF (Decision Tree Scan with Union-Find) не применим для сжатого
изображения, так как использует свойства 8-связности, которыми наша новая
связность не обладает.
}

\texten{
As before, the goal is to reduce the number of pixel lookup and label copy operations.
This is achieved with careful consideration of only the necessary neighbors,
stopping once the set of one or two label operations that are needed is
determined. The decision tree for processing an object pixel is
shown on \autoref{fig:dtsuf}.
}\textru{
Как и раньше, цель состоит в том, чтобы уменьшить число запросов на содержание
пикселя и копирований меток.
Это достигается аккуратным рассмотрением соседей в оптимальном порядке и
остановкой рассмотрения, как только достаточная для выполнения всех необоходимых
действий информация была получена.
Смотри дерево на \autoref{fig:dtsuf}.
}
